---
title: "Introduction to {crane}"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is {crane}?

The {crane} package supplements the {gtsummary} package to ease the process of creating Roche clinical trial tables. 
{gtsummary} is a general framework for building tables, and {crane} utilizes the frame work and exports functions meant to simplify the creation of summary tables in pharmaceutical industry.

The {crane} package includes wrapper functions for {gtsummary} functions to make it easier to apply common pharma-specific modifications to table formats.
For example, demographics tables which use certain variables and formatting conventions as a default. 

When the crane package is loaded into your R session, a couple of things occur.

- As expected, the {crane} package is loaded, but the {gtsummary}, {cards}, and {cardx} packages are also loaded.
- The Roche gtsummary theme is set by running `crane::theme_gtsummary_roche()`. This theme changes a few default behaviors for {gtsummary} tables; for example, all p-values are rounded with `label_roche_pvalue()`, and tables are printed with {flextable} instead of {gt}. See the function's help file for details.

```{r}
library(crane)
theme_gtsummary_roche()
```

#### When should I add a new function to  {crane}?

While all Roche tables can be created using {gtsummary}, for some tables the code can become long.
In these cases, adding a new function to {crane} is warranted.

For example, our standard demographics table can be created with `gtsummary::tbl_summary()`.
But for such a common table, the code is somewhat long, and we therefore created `crane::tbl_roche_summary()`.

To create the Roche standard demographics table, we simple call:

```{r}
cards::ADSL |>
  tbl_roche_summary(
    by = ARM,
    include = c(AGE, RACE),
    nonmissing = "always"
  )
```

To create this table with `gtsummary::tbl_summary()`, we'd need the following code.
As some of these steps are not the most straightforward, we created the `tbl_roche_summary()` wrapper.

```r
cards::ADSL |> 
  tbl_summary(
    by = ARM, 
    include = c(AGE, RACE),
    type = list(AGE = "continuous2"),
    statistic = list(AGE = c("{mean} ({sd})", "{median}", "{min} - {max}")),
    digits = all_categorical() ~ list(p = 1),
    missing = "always",
    missing_stat = "{N_nonmiss}",
    missing_text = "n"
  ) |>
    # remove default footnote
    remove_footnote_header(columns = everything()) |>
    # remove the default "Characteristic" header
    modify_header(label = "") |>
    # convert "0 (0.0%)" to "0"
    gtsummary::modify_post_fmt_fun(
      fmt_fun = ~ifelse(. == "0 (0.0%)", "0", .),
      columns = all_stat_cols()
    ) |>
    # sort the missing row to just below the header row
    modify_table_body(
      function(x) {
        x |>
          dplyr::mutate(
            .by = "variable",
            dplyr::across(dplyr::everything(), ~ .[order(row_type != "label", row_type != "missing")])
          )
      }
    )
```

Another case where we may create helper functions in {crane}, is when the table can only be created in {gtsummary} using an ARD-first approach.
Generally, creating an ARD of all results in a summary table, then creating then table is more cumbersome.
In these cumbersome cases, a new function in {crane} is warranted.

## Regression Tables

In addition to summary tables, {crane} provides convenient wrappers for creating regression tables from various model types. These functions ensure proper integration with {gtsummary} and provide model-specific validation.

### MMRM Regression Tables

The `tbl_mmrm()` function provides a convenient wrapper for creating regression tables from Mixed Models for Repeated Measures (MMRM). This function works seamlessly with the {mmrm} package and {gtsummary}.

```r
library(mmrm)

# Fit an MMRM model
fit_mmrm <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = mmrm::fev_data
)

# Create a regression table
tbl_mmrm(fit_mmrm)
```

The function is essentially a wrapper around `gtsummary::tbl_regression()` with additional validation for mmrm objects. All arguments from `tbl_regression()` can be passed through:

```r
# Customize the output
tbl_mmrm(
  fit_mmrm,
  include = c("RACE", "SEX", "ARMCD"),
  conf.level = 0.90
)
```


