---
title: "Modular TLG Pipeline"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

{crane} separates TLG creation into three independent layers:

1. **Compute** — produce a statistics data frame
2. **Tabulate** — render it as a `{gtsummary}` table
3. **Customize** — adjust formatting, headers, or styling

Each layer can be swapped or modified without touching the others.
This article walks through the pattern using `tbl_coxph()` as the primary
example, then shows how the same idea applies to other crane components.

## The three layers

### Layer 1: Compute

`get_cox_pairwise_df()` fits Cox models and returns a plain data frame.
No formatting decisions are made here beyond rounding HR and CI to two
decimal places.

```{r compute, message=FALSE}
library(crane)
library(survival)

surv_data <- survival::lung |>
  dplyr::mutate(
    arm = factor(
      sample(c("A", "B", "C"), dplyr::n(), replace = TRUE),
      levels = c("A", "B", "C")
    )
  )

pairwise_df <- get_cox_pairwise_df(
  data = surv_data,
  formula = Surv(time, status) ~ arm,
  arm = "arm",
  ref_group = "A"
)

pairwise_df
```

The result is a regular `data.frame` with one row per comparison arm.
You can inspect it, filter it, or replace it entirely with results from
a different statistical engine.

### Layer 2: Tabulate

`tbl_coxph()` takes the data frame and renders it as a `{gtsummary}` table.
It does not re-compute anything — it only formats and displays.

```{r tabulate}
tbl_coxph(pairwise_df)
```

### Layer 3: Customize

Formatting can be adjusted at multiple points.

**Before tabulation** — pre-format columns in the data frame.
Character values in the p-value column are passed through as-is:

```{r customize-pre}
pval_col <- grep("p-value", names(pairwise_df), value = TRUE)
custom_df <- pairwise_df
custom_df[[pval_col]] <- ifelse(
  custom_df[[pval_col]] < 0.001, "<0.001",
  sprintf("%.3f", custom_df[[pval_col]])
)

tbl_coxph(custom_df)
```

**After tabulation** — use `{gtsummary}` modifiers on the output:

```{r customize-post}
tbl_coxph(pairwise_df) |>
  gtsummary::modify_header(label ~ "**Comparison**")
```

**Selecting columns** — pass a subset of the data frame to control which
statistics appear:

```{r customize-cols}
# HR and CI only (no p-value)
tbl_coxph(pairwise_df[, c("HR", "95% CI")])
```

## Where digits are set

| What | Where | Default | How to change |
|---|---|---|---|
| HR | `get_cox_pairwise_df()` | `%.2f` | Modify the data frame column |
| 95% CI | `get_cox_pairwise_df()` | `(%.2f, %.2f)` | Modify the data frame column |
| p-value | `.format_pvalue()` in `tbl_coxph()` | `%.4f` / `<0.0001` | Pre-format as character |

## Other modular components

The compute → tabulate → customize pattern appears throughout {crane}:

### `tbl_hierarchical_rate_by_grade()` + `add_grade_column()`

The grade-label column is injected *after* table construction by
`add_grade_column()`. This avoids a Cartesian join when merging tables
with `tbl_merge()`:

```r
# Compute + tabulate
tbl <- tbl_hierarchical_rate_by_grade(data, ...)

# Customize (add grade column after merging)
tbl_merge(list(tbl_a, tbl_b)) |>
  add_grade_column()
```

### `tbl_roche_subgroups()` + `add_forest()`

The forest plot is a separate post-processing step. The subgroup table
is built first, then the plot is layered on:

```r
tbl <- data |>
  tbl_roche_subgroups(
    rsp = "is_event",
    by = all_of(arm_var),
    subgroups = c("SEX", "RACE"),
    .tbl_fun = my_regression_fun
  )

# Add forest plot as a separate layer
plt <- tbl |> add_forest()
```

### `tbl_with_pools()`

A pooling wrapper that accepts *any* `.tbl_fun`. The pooling logic
(combining arms into summary groups) is independent of the table type:

```r
tbl_with_pools(
  data = adae,
  pools = list("Xanomeline" = c("Xanomeline Low Dose", "Xanomeline High Dose")),
  by = "TRTA",
  .tbl_fun = tbl_hierarchical
)
```

## Swapping layers

Because each layer is independent, you can swap any of them:
 
- **Bring your own data frame** — compute statistics in SAS, Python, or
  another R package, then pass the result to `tbl_coxph()`.
- **Use a different `.tbl_fun`** — `tbl_roche_subgroups()` accepts any
  function that returns a `gtsummary` object. Swap `tbl_regression()` for
  a custom logistic regression wrapper.
- **Post-process with standard gtsummary tools** — `modify_header()`,
  `modify_fmt_fun()`, `modify_footnote()`, `modify_spanning_header()` all
  work on crane output because it's standard `gtsummary` under the hood.
