---
title: "Working with Pooled Treatment Arms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Pooled Treatment Arms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In clinical reporting, it is incredibly common to require tables that display both the original treatment arms (e.g., "Dose A", "Dose B") alongside pooled arms (e.g., "Any Dose") or a total column ("All Patients").

Because these groups overlap, a single patient must technically be counted in multiple columns. Modifying the underlying ADaM datasets to achieve this introduces duplicated `USUBJID`s, which breaks standard CDISC data integrity and can cause severe downstream calculation errors if not managed perfectly.

`{crane}` provides two ways to handle this. We **highly recommend Method 1** for all standard TLG generation.

## Method 1: The Recommended Approach (tbl_with_pools)

For any table generated using `{gtsummary}` or custom extensions (like `tbl_hierarchical_rate_and_count`), you should use `tbl_with_pools()`.

### Why is this the recommended approach?

- **Pristine Data Integrity:** Your ADaM datasets remain untouched in your environment. Patients are never duplicated.

- **Automatic Denominator Management:** It safely subsets both your analysis dataset and your denominator dataset simultaneously.

- **Safe Totals:** Bypasses the need for `gtsummary::add_overall()`, which would mathematically double-count patients if applied to a pre-pooled dataset.

### Setup Dummy Data

```{r}
library(dplyr)
library(gtsummary)
library(crane)

# Minimal ADSL and ADAE datasets
adsl <- data.frame(
  USUBJID = c("001", "002", "003", "004", "005"),
  TRT01A = c("Drug A", "Drug A", "Drug B", "Drug C", "Drug C"),
  AGE = c(45, 50, 60, 65, 55),
  stringsAsFactors = FALSE
)

adae <- data.frame(
  USUBJID = c("001", "001", "002", "004", "005"),
  TRT01A = c("Drug A", "Drug A", "Drug A", "Drug C", "Drug C"),
  AEBODSYS = c("SOC1", "SOC1", "SOC2", "SOC1", "SOC2"),
  stringsAsFactors = FALSE
)
```

### Defining the Pools

Define your pools using a simple named list. Use the `"all"` keyword to generate a total column.

```{r}
my_pools <- list(
  "Drug A + B"   = c("Drug A", "Drug B"),
  "All Patients" = "all"
)
```

### Example: Efficacy/Demographics (No Denominator)

Pass your data directly into `tbl_with_pools()`, specifying your target gtsummary function via `.tbl_fun`. All other arguments are passed straight through to the target function.

```{r}
tbl_demo <- tbl_with_pools(
  data = adsl,
  pools = my_pools,
  arm_var = "TRT01A",
  denominator = NULL,        # tbl_summary does not use a denominator
  keep_original = TRUE,      # Keep Drug A, Drug B, and Drug C columns
  .tbl_fun = tbl_summary,
  include = AGE              # Argument passed to tbl_summary
)

tbl_demo
```

### Example: Safety (With Denominator)

For safety tables that require a denominator, simply pass adsl to the denominator argument. The function handles the subsetting geometry for you.

```{r}
tbl_ae <- tbl_with_pools(
  data = adae,
  pools = my_pools,
  arm_var = "TRT01A",
  denominator = adsl,
  keep_original = TRUE,
  .tbl_fun = tbl_hierarchical_rate_and_count,
  variables = c(AEBODSYS)
)

tbl_ae
```

## Method 2: The Pre-processing Fallback (`df_add_poolings`)
⚠️ _DISCLAIMER: This is a risky function. Please consider using `tbl_with_pools()` instead._

Sometimes, you need pooled data outside of `gtsummary`—for instance, when generating a custom `ggplot2` graphic, exporting a flattened `.csv` of grouped statistics, or using an alternative reporting engine.

In these specific scenarios, you can use `df_add_poolings()` to physically append the pooled rows directly onto your data frames.

### The Danger of the "Total" Column

If you set `keep_original = TRUE` (keeping your base arms) and add a pool of `"all"`, **your dataset will double in size**. If a downstream function attempts to calculate a "Total" using this newly expanded data, it will count every patient twice.

The function will explicitly warn you when you do this:

```{r}
adam_db <- list(adsl = adsl, adae = adae)

adam_db_pooled <- df_add_poolings(
  adam_db = adam_db,
  pools = my_pools, 
  arm_var = "TRT01A",
  keep_original = TRUE
)
```
### Safe Usage

When using `df_add_poolings()`, it is often safest to set `keep_original = FALSE`. This replaces the original arms entirely with your new pools, guaranteeing that no single patient exists in the dataset more than once per pooled definition.

```{r}
adam_db_safe <- df_add_poolings(
  adam_db = adam_db,
  pools = list("Drug A + B" = c("Drug A", "Drug B")), 
  arm_var = "TRT01A",
  keep_original = FALSE
)

# Observe that TRT01A now only contains "Drug A + B" for the applicable rows
table(adam_db_safe$adsl$TRT01A)
```

## Tip: Using External Configuration (YAML)
To keep your template scripts clean and generic across studies, both of these functions are designed to accept configurations directly from a YAML file (e.g., `arg_specs.yaml`).

`arg_specs.yaml`:

```{yaml}
poolings_config:
  keep_original: TRUE
  arm_var: "TRT01A"
  pools:
    "Drug A + B": ["Drug A", "Drug B"]
    "Total": ["all"]
Inside your script:
```

```{r}
pool_args <- yaml::read_yaml("arg_specs.yaml")$poolings_config

# Pass cleanly into the wrapper
tbl <- tbl_with_pools(
  data          = adae,
  pools         = pool_args$pools,
  arm_var       = pool_args$arm_var,
  denominator   = adsl,
  keep_original = pool_args$keep_original,
  .tbl_fun      = tbl_summary
)
```
