---
title: "Working with Pooled Treatment Arms"
output: rmarkdown::html_vignette
vignette: >
  %\\VignetteIndexEntry{Working with Pooled Treatment Arms}
  %\\VignetteEngine{knitr::rmarkdown}
  %\\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In clinical reporting, it is incredibly common to require tables that display both the original treatment arms (e.g., "Dose A", "Dose B") alongside pooled arms (e.g., "Any Dose") or a total column ("All Patients"). You may also need to define highly specific cohorts based on clinical flags.

Because these groups overlap, a single patient must technically be counted in multiple columns. Modifying the underlying ADaM datasets to achieve this introduces duplicated `USUBJID`s, which breaks standard CDISC data integrity and can cause severe downstream calculation errors if not managed perfectly.

`{crane}` provides two ways to handle this. We **highly recommend Method 1** for all standard TLG generation.

## Method 1: The Recommended Approach (tbl_with_pools)

For any table generated using `{gtsummary}` or custom extensions (like `tbl_hierarchical_rate_and_count`), you should use `tbl_with_pools()`.

### Why is this the recommended approach?

- **Pristine Data Integrity:** Your ADaM datasets remain untouched in your environment. Patients are never duplicated.

- **Automatic Denominator Management:** It safely subsets both your analysis dataset and your denominator dataset simultaneously, and dynamically evaluates complex logical expressions across both.

- **Safe Totals:** Bypasses the need for `gtsummary::add_overall()`, which would mathematically double-count patients if applied to a pre-pooled dataset.

### Setup Dummy Data

Notice we are including standard analysis flags (`AFL`, `BFL`) which we will use later to define complex pools.

```{r}
library(dplyr)
library(gtsummary)
library(rlang)
library(crane)

# Minimal ADSL and ADAE datasets
adsl <- data.frame(
  USUBJID = c("001", "002", "003", "004", "005"),
  TRT02A = c("Drug A", "Drug A", "Drug B", "Drug C", "Drug C"),
  AGE = c(45, 50, 60, 65, 55),
  AFL = c("Y", "Y", "N", "Y", "N"),
  BFL = c("Y", "Y", "Y", "N", "Y"),
  stringsAsFactors = FALSE
)

adae <- data.frame(
  USUBJID = c("001", "001", "002", "004", "005"),
  TRT02A = c("Drug A", "Drug A", "Drug A", "Drug C", "Drug C"),
  AEBODSYS = c("SOC1", "SOC1", "SOC2", "SOC1", "SOC2"),
  AEDECOD = c("PT1", "PT2", "PT3", "PT1", "PT4"),
  AFL = c("Y", "Y", "Y", "Y", "N"),
  BFL = c("Y", "Y", "Y", "N", "Y"),
  stringsAsFactors = FALSE
)
```

### Defining the Pools

`{crane}` supports a mixed-definition pooling list. You can seamlessly combine:
1. **Character vectors:** For standard string matching on the arm variable.
2. **The `"all"` keyword:** To create an absolute total column.
3. **Logical Expressions (`rlang::expr()`):** To define complex, dynamic cohorts based on multiple variables.

```{r}
my_pools <- list(
  "Complex Pool (Flags)" = expr(AFL == "Y" & BFL == "Y"),
  "Drug A + B"           = c("Drug A", "Drug B"),
  "All Patients"         = "all"
)
```

### Example: Efficacy/Demographics (No Denominator)

Pass your data directly into `tbl_with_pools()`, specifying your target gtsummary function via `.tbl_fun`. All other arguments are passed straight through to the target function.

```{r}
tbl_demo <- tbl_with_pools(
  data = adsl,
  pools = my_pools,
  by = "TRT02A",
  denominator = NULL, # tbl_summary does not use a denominator
  keep_original = TRUE, # Keep Drug A, Drug B, and Drug C columns
  .tbl_fun = tbl_summary,
  include = AGE # Argument passed to tbl_summary
)

tbl_demo
```

### Example: Safety (With Denominator)

For safety tables that require a denominator, simply pass adsl to the denominator argument. The function handles the subsetting geometry for you.

```{r}
tbl_ae <- tbl_with_pools(
  data = adae,
  pools = my_pools,
  by = "TRT02A",
  denominator = adsl,
  keep_original = TRUE,
  .tbl_fun = tbl_hierarchical_rate_and_count,
  variables = c(AEBODSYS, AEDECOD)
)

tbl_ae
```

## Method 2: The Pre-processing Fallback (`df_add_poolings`)
_**DISCLAIMER**: This is a risky function. Please consider using `tbl_with_pools()` instead._

Sometimes, you need pooled data outside of `gtsummary`—for instance, when generating a custom `ggplot2` graphic, exporting a flattened `.csv` of grouped statistics, or using an alternative reporting engine.

In these specific scenarios, you can use `df_add_poolings()` to physically append the pooled rows directly onto your data frames.

### The Danger of the "Total" Column

If you set `keep_original = TRUE` (keeping your base arms) and add a pool of `"all"`, **your dataset will double in size**. If a downstream function attempts to calculate a "Total" using this newly expanded data, it will count every patient twice.

The function will explicitly warn you when you do this:

```{r}
adam_db <- list(adsl = adsl, adae = adae)

adam_db_pooled <- df_add_poolings(
  adam_db = adam_db,
  pools = my_pools,
  arm_var = "TRT02A",
  keep_original = TRUE
)
```

### Safe Usage

When using `df_add_poolings()`, it is often safest to set `keep_original = FALSE`. This replaces the original arms entirely with your new pools, guaranteeing that no single patient exists in the dataset more than once per pooled definition.

```{r}
adam_db_safe <- df_add_poolings(
  adam_db = adam_db,
  pools = list(
    "Drug A + B" = c("Drug A", "Drug B"),
    "ccRCC Patients" = expr(AFL == "Y" & BFL == "Y")
  ),
  arm_var = "TRT02A",
  keep_original = FALSE
)

# Observe that TRT02A now only contains our specific pools for the applicable rows
table(adam_db_safe$adsl$TRT02A)
```

## Tip: Using External Configuration (`YAML`)
To keep your template scripts clean and generic across studies, both of these functions are designed to accept configurations directly from a `YAML` file (e.g., `meta_rules.yaml`).

*Note: Because YAML cannot natively store R language objects, complex conditions will be read into R as plain text strings. However, you can easily convert these strings back into evaluable code by passing them through `rlang::parse_expr()` after importing your YAML configuration.*

`meta_rules.yaml`:

```{yaml, eval = FALSE}
poolings_config:
  keep_original: FALSE
  arm_var: "TRT02A"
  pools:
    "Drug A + B": ["Drug A", "Drug B"]
    "Total": ["all"]
    # Write the expression as a plain string
    "AB Patients": "AFL == 'Y' & BFL == 'Y'"
```

Inside your script:

```{r, eval = FALSE}
library(rlang)

# 1. Read the YAML
pool_args <- yaml::read_yaml("meta_rules.yaml")$poolings_config

# 2. Convert the string into a parsed expression
pool_args$pools$`AB Patients` <- rlang::parse_expr(pool_args$pools$`AB Patients`)

# 3. Pass it to your function! (is.language() will now correctly evaluate to TRUE)
tbl <- tbl_with_pools(
  data          = adae,
  pools         = pool_args$pools,
  by            = pool_args$arm_var,
  denominator   = adsl,
  keep_original = pool_args$keep_original,
  .tbl_fun      = tbl_summary
)
```
